Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)

### 2017-09-21 code/Rmd/ALL_REPROCESSED.Rmd

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c('367_120502')

@@

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

  # Get design matrix
  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]

  # Estimate voom weights
  voomWeights = limma::voom(cnt[rownames(lcpm),colnames(lcpm)], design = designMat, plot=F)$weights

  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
                              .combine = rbind,
                              .packages = c('lme4', 'dplyr'),
                              .export = c('voomWeights')) %dopar% {

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             covars[rownames(designMat),groupingVar,drop=F])
                                formula = paste(setdiff(colnames(data), c('Gx', groupingVar)), collapse = '+') %>%
                                  paste('Gx ~',.) %>%
                                  c(.,paste0('(1|',groupingVar,')')) %>%
                                  paste(collapse = '+')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(Gene.ID = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           c('individualIdentifier'), 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

                            return(list(modelBIC))
                          }
    allModelBIC = plyr::join_all(allModelBIC)

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -Gene.ID) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = (better/(better+worse)) * 100)

    print(nChange)

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > 1) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(Gene.ID, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('BrainRegion.Diagnosis')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Study', 'Gender', 'PMI', 'AOD', 'RIN', 'RIN2')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('Batch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`

### Normalisation
1. BrainRegion.Diagnosis is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

@@

### Residual calculation
Calculate weighted residuals and add back "BrainRegion.Diagnosis" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

### Differential expression analysis (with BrainRegion.Diagnosis as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp1, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, adjust.covars, drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

@@

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

@@

### Differential expression analysis (with BrainRegion.ApoE as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.ApoE', setdiff(adjust.covars, 'BrainRegion.Diagnosis')), drop = F], 
                         Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

@@

# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(adjust.covars, c('BrainRegion.Diagnosis', 'Gender'))), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(setdiff(adjust.covars, 'AOD')), drop = F], Intercept = F)
ind = grep('BrainRegion.Diagnosis', colnames(DESIGN$design))
DESIGN$design[,ind] = DESIGN$design[,ind]*COVARIATES$AOD
colnames(DESIGN$design)[ind] = paste(colnames(DESIGN$design)[ind], 'AOD', sep = '_')
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp4.1, fig.height=10, fig.width=15}
# Cell type markers
CELL.FRAC = downloadFile('syn10163089')
all.used.ids = c(all.used.ids, 'syn10163089')

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

@@

# Download all related genesets from synapse
load(synGet('syn4867851')@filePath)
all.used.ids = c(all.used.ids, 'syn4867851')
geneSet.ref = GeneSets[c('KEGG_2016', 'GO_Biological_Process_2015', 'Reactome_2016', 'WikiPathways_2016')] %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

@@

>>>>

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c('367_120502', '1950_TCX', '1950_CER', '1925_TCX', '1957_CER')

@@

### Normalisation (iterative design)
1. BrainRegion.Diagnosis is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r iterative.norm, results='asis'}
# Primary variable of interest
primaryVariable = c('BrainRegion.Diagnosis');
notEstimable = c('BrainRegion','Diagnosis','BrainRegion.ApoE')
postAdjustCovars = c()

# Assign residual covariates
residualCovars = setdiff(preAdjustedSigCovars$significantCovars, 
                         c(postAdjustCovars, primaryVariable, notEstimable, 'individualIdentifier'))
residualSigCovars = preAdjustedSigCovars
covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and individualIdentifier as random effect'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$individualIdentifier,
                                             method = 'lmer')

  if (!is.nan(correlation$cor)){
    # Re-estimate voom weights
    VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                                design=DM1$design, plot=F, 
                                block = COVARIATES$individualIdentifier, 
                                correlation = correlation$cor)

    # Fit linear model using new weights and new design
    VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
                              design = DM1$design,
                              weights = VOOM.GENE_EXPRESSION$weights,
                              block = COVARIATES$individualIdentifier, 
                              correlation = correlation$cor)

    # Residuals after normalisation
    RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                  CQN.GENE_EXPRESSION$E)
  } else {
    notEstimable = c(notEstimable, postAdjustCovars[length(postAdjustCovars)])
    postAdjustCovars = postAdjustCovars[1:(length(postAdjustCovars)-1)]
  }

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates),
                         c(postAdjustCovars, primaryVariable, 'COVARIATES$individualIdentifier', notEstimable))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, primaryVariable,
                                                               'individualIdentifier', notEstimable))
  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```

@@

### Residual calculation
Calculate weighted residuals and add back "BrainRegion.Diagnosis" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(VOOM.ADJUSTED.FIT$design[,varsToAddIn])
```

@@

### Differential expression analysis (with BrainRegion.Diagnosis as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp1, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c("BrainRegion.Diagnosis", postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

@@

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

@@

### Differential expression analysis (with BrainRegion.ApoE as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.ApoE', setdiff(postAdjustCovars, 'BrainRegion.Diagnosis')), drop = F],
                         Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

@@

# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(postAdjustCovars, c('BrainRegion.Diagnosis', 'Gender'))), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp4, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis', setdiff(postAdjustCovars, 'AOD')), drop = F], Intercept = F)
ind = grep('BrainRegion.Diagnosis', colnames(DESIGN$design))
DESIGN$design[,ind] = DESIGN$design[,ind]*COVARIATES$AOD
colnames(DESIGN$design)[ind] = paste(colnames(DESIGN$design)[ind], 'AOD', sep = '_')
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp5, fig.height=10, fig.width=15, eval = F}
# Cell type markers
CELL.FRAC = downloadFile('syn10163089')
all.used.ids = c(all.used.ids, 'syn10163089')

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

@@

# Download all related genesets from synapse
load(synGet('syn4867851')@filePath)
all.used.ids = c(all.used.ids, 'syn4867851')
geneSet.ref = GeneSets[c('KEGG_2016', 'GO_Biological_Process_2015', 'Reactome_2016', 'WikiPathways_2016')] %>%
  lapply(function(x){
    lapply(x, function(y){
      sapply(y, function(y) str_split(y, '\\,')[[1]][1])
    })
  }) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

@@

# Store all model fits
save(list = 'all.fit', file = 'AMP_allModelFits.RData')
AFIT_OBJ = File('AMP_allModelFits.RData',
                name = 'All Model Fit', 
                parentId = CODE$properties$id)
annotations(AFIT_OBJ) = annotations(CODE)
annotations(AFIT_OBJ)$dataSubType = 'allModelFit'
AFIT_OBJ = synStore(AFIT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
synChangeEntityACL(AFIT_OBJ@properties$id, new.racl)
